How many speech sounds are there in the world's languages?

### 2023-06-06 paper.Rmd
edits

How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasingly new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases. Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires always yield more phoneme types cross-linguistically. As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law. We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5x more than what is currently documented in the world's languages (~35%). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.

@@

To address these issues, we modeled the distributional properties of documented phonemes. First, we found that frequencies of phoneme types in existing cross-linguistic databases are well described by Zipf's law, regardless of whether samples were controlled for areal and genetic biases. Systems that obey Zipf's law, under mild assumptions, also follow Herdan-Heaps' law [@BaezaNavarro2000;@vanLeijenhorst2005], which predicts type count from the number of tokens with only two tunable parameters. Herdan-Heaps' law was originally intended to describe how vocabulary size increases as more words are observed in a document or set of documents. Extending this approach, we demonstrate that the number of phoneme types in a sample of languages also grows according to Herdan-Heaps' law. To our knowledge, we are the first to apply Herdan-Heaps' law to the domain of segmental phonology by reformulating it as estimator of the number of phonemes in groups of languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by applying an additional and formally unrelated tool borrowed from Ecology, which we refer to as the Chao-Shen estimator. The predictions of this estimator confirm that the relationship between sample size and segment type count is best described by a curve that approximates Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did not converge on any single number of phoneme types as number of languages increased.

Assuming a target of 7151 known languages,^[https://www.ethnologue.com/guides/how-many-languages?] our models predict a range of total segment types from 4592 to 4850 (mean = 4721). Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183. Therefore, the models argue for the existence of approximately 1500 living but undocumented phoneme types. As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K. The Chao-Shen model produced much more conservative estimates for samples of size 10K and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100K languages or more. However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value. Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps.

@@

However, most widely accepted feature sets contain features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root). Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language. And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues. Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors ($2^24$). Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature. Making finer and finer contrasts may always produce more and more features. Thus, the exponent in the Feature Bounding calculation can itself (by hypothesis) be infinite, in line with the predictions of the models presented here, and therefore with the notion of infinite possible phonemes. That said, contemporary feature sets can capture phonological classes for place and manner of articulation, which neatly classify the phonological patterns and processes in and across languages (cf. @Mielke2008).

@@

Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may be. Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and the cross-linguistic study of languages.

@@

Steven Moran was funded by the Swiss National Science Foundation (Grant No. PCEFP1_186841).

>>>>

How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasingly new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases. Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires cross-linguistically always yields more phoneme types. As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law. We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5x more than what is currently documented in the world's languages (~35%). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.

@@

To address these issues, we modeled the distributional properties of documented phonemes. First, we found that frequencies of phoneme types in existing cross-linguistic databases are well described by Zipf's law, regardless of whether samples were controlled for areal and genetic biases. Systems that obey Zipf's law, under mild assumptions, also follow Herdan-Heaps' law [@BaezaNavarro2000;@vanLeijenhorst2005], which predicts type count from the number of tokens with only two tunable parameters. Herdan-Heaps' law was originally intended to describe how vocabulary size increases as more words are observed in a document or set of documents. Extending this approach, we demonstrate that the number of phoneme types in a sample of languages also grows according to Herdan-Heaps' law. To our knowledge, we are the first to apply Herdan-Heaps' law to the domain of segmental phonology by reformulating it as estimator of the number of phonemes in groups of languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by applying an additional and formally unrelated tool borrowed from ecology, which we refer to as the Chao-Shen estimator. The predictions of this estimator confirm that the relationship between sample size and segment type count is best described by a curve that approximates Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did not converge on any single number of phoneme types as number of languages increased.

Assuming a target of 7151 known languages,^[https://www.ethnologue.com/guides/how-many-languages?] our models predict a range of total segment types from 4592 to 4850 (mean = 4721). Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183. Therefore, the models argue for the existence of approximately 1500 yet to be documented phoneme types. As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K. The Chao-Shen model produced much more conservative estimates for samples of size 10K and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100K languages or more. However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value. Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps' law.

@@

However, most widely accepted feature sets contain features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root). Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language. And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues. Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24). Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature. Making finer and finer contrasts may always produce more and more features. Thus, the exponent in the Feature Bounding calculation can itself (by hypothesis) be infinite, in line with the predictions of the models presented here, and therefore with the notion of infinite possible phonemes. That said, contemporary feature sets can capture phonological classes for place and manner of articulation, which neatly classify the phonological patterns and processes in and across languages (cf. @Mielke2008).

@@

Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may or may not be. Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.

@@

SM was funded by the Swiss National Science Foundation (Grant No. PCEFP1_186841).

# Author contributions

SM conceptualization. SM and NAL designed the study, interpreted the results, and wrote the paper.

### 2023-06-06 paper.tex
edits

How many speech sounds are there in the world's languages? To answer
this question, we extend two formal models of lexical diversity to
segmental phonology. Zipf's law states that most word types are rare;
Herdan-Heaps' law states that vocabulary size grows indefinitely with
increasingly new text. We find that both models produce accurate
estimates of the diversity of contrastive speech sounds (phonemes) in
existing typological databases. Our findings suggest that (a) most
phonemes are difficult to detect because they are rare, and (b)
documenting more phonological repertoires always yield more phoneme
types cross-linguistically. As a sanity check, we apply an unrelated
estimator of species diversity from biology, Chao-Shen, and show that
its estimates align with Herdan-Heaps' law. We use both models to
extrapolate phoneme counts to undocumented languages, resulting in an
estimate of 4600-4900 distinct phonemes in the full set of languages
thought to be spoken today (\textasciitilde7-8k), which is roughly 1.5x
more than what is currently documented in the world's languages
(\textasciitilde35\%). By extrapolating to hypothetical sample sizes,
e.g., the number of languages estimated to have been spoken through
time, and beyond, we find no indication of convergence: infinite
languages mean infinite phonemes.

@@

To address these issues, we modeled the distributional properties of
documented phonemes. First, we found that frequencies of phoneme types
in existing cross-linguistic databases are well described by Zipf's law,
regardless of whether samples were controlled for areal and genetic
biases. Systems that obey Zipf's law, under mild assumptions, also
follow Herdan-Heaps' law
(\protect\hyperlink{ref-BaezaNavarro2000}{Baeza--Yates \& Navarro 2000};
\protect\hyperlink{ref-vanLeijenhorst2005}{Leijenhorst \& Van der Weide
2005}), which predicts type count from the number of tokens with only
two tunable parameters. Herdan-Heaps' law was originally intended to
describe how vocabulary size increases as more words are observed in a
document or set of documents. Extending this approach, we demonstrate
that the number of phoneme types in a sample of languages also grows
according to Herdan-Heaps' law. To our knowledge, we are the first to
apply Herdan-Heaps' law to the domain of segmental phonology by
reformulating it as estimator of the number of phonemes in groups of
languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by
applying an additional and formally unrelated tool borrowed from
Ecology, which we refer to as the Chao-Shen estimator. The predictions
of this estimator confirm that the relationship between sample size and
segment type count is best described by a curve that approximates
Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did
not converge on any single number of phoneme types as number of
languages increased.

Assuming a target of 7151 known languages,\footnote{\url{https://www.ethnologue.com/guides/how-many-languages}?}
our models predict a range of total segment types from 4592 to 4850
(mean = 4721). Based on PHOIBLE, our current best guess of the diversity
of documented phoneme types is 3183. Therefore, the models argue for the
existence of approximately 1500 living but undocumented phoneme types.
As we probe increasingly larger and theoretically motivated language
sample sizes, i.e., languages that existed \emph{through time} (e.g.,
estimates including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel
2000}) or 64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})),
the Herdan-Heaps' model predicted a continual increase in the number of
unique phonemes, reaching values as high as 19K. The Chao-Shen model
produced much more conservative estimates for samples of size 10K and
above, even appearing to settle on an asymptote of roughly 5500 phonemes
for samples of 100K languages or more. However, when we compare the
predictions of Herdan-Heaps and Chao-Shen against known type counts in a
closed dataset, the source of the discrepancy emerged: as the sample
grows, Chao-Shen lags behind Herdan-Heaps in approximating the true
value. Thus, given a larger pool of training data, we expect Chao-Shen
to approximate more closely the estimates of Herdan-Heaps.

@@

However, most widely accepted feature sets contain features that are
specified at different levels of phonetic detail: some features are more
abstract (i.e., general similarities such as +/- syllabic), while others
are more concrete (i.e., closer to the articulatory reality, e.g.,
advanced tongue root). Some are grounded in pure articulatory or
acoustic phonetics, while others involve phonological properties of
language. And looming in the background of these discussions, the debate
of whether distinctive features are innate and universal or rather
learned and emergent continues. Nevertheless, in general, distinctive
feature theories of different theoretical stances propose around two
dozen features (\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume
2006}), which still has the potential to generate over 16 million
distinct feature vectors (\(2^24\)). Current consensus aside, there does
not appear to be a principled limit to the degree of phonetic
specificity that can be encoded by a feature. Making finer and finer
contrasts may always produce more and more features. Thus, the exponent
in the Feature Bounding calculation can itself (by hypothesis) be
infinite, in line with the predictions of the models presented here, and
therefore with the notion of infinite possible phonemes. That said,
contemporary feature sets can capture phonological classes for place and
manner of articulation, which neatly classify the phonological patterns
and processes in and across languages (cf. Mielke
(\protect\hyperlink{ref-Mielke2008}{2008})).

@@

Steven Moran was funded by the Swiss National Science Foundation (Grant
No.~PCEFP1\_186841).

>>>>

How many speech sounds are there in the world's languages? To answer
this question, we extend two formal models of lexical diversity to
segmental phonology. Zipf's law states that most word types are rare;
Herdan-Heaps' law states that vocabulary size grows indefinitely with
increasingly new text. We find that both models produce accurate
estimates of the diversity of contrastive speech sounds (phonemes) in
existing typological databases. Our findings suggest that (a) most
phonemes are difficult to detect because they are rare, and (b)
documenting more phonological repertoires cross-linguistically always
yields more phoneme types. As a sanity check, we apply an unrelated
estimator of species diversity from biology, Chao-Shen, and show that
its estimates align with Herdan-Heaps' law. We use both models to
extrapolate phoneme counts to undocumented languages, resulting in an
estimate of 4600-4900 distinct phonemes in the full set of languages
thought to be spoken today (\textasciitilde7-8k), which is roughly 1.5x
more than what is currently documented in the world's languages
(\textasciitilde35\%). By extrapolating to hypothetical sample sizes,
e.g., the number of languages estimated to have been spoken through
time, and beyond, we find no indication of convergence: infinite
languages mean infinite phonemes.

@@

To address these issues, we modeled the distributional properties of
documented phonemes. First, we found that frequencies of phoneme types
in existing cross-linguistic databases are well described by Zipf's law,
regardless of whether samples were controlled for areal and genetic
biases. Systems that obey Zipf's law, under mild assumptions, also
follow Herdan-Heaps' law
(\protect\hyperlink{ref-BaezaNavarro2000}{Baeza--Yates \& Navarro 2000};
\protect\hyperlink{ref-vanLeijenhorst2005}{Leijenhorst \& Van der Weide
2005}), which predicts type count from the number of tokens with only
two tunable parameters. Herdan-Heaps' law was originally intended to
describe how vocabulary size increases as more words are observed in a
document or set of documents. Extending this approach, we demonstrate
that the number of phoneme types in a sample of languages also grows
according to Herdan-Heaps' law. To our knowledge, we are the first to
apply Herdan-Heaps' law to the domain of segmental phonology by
reformulating it as estimator of the number of phonemes in groups of
languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by
applying an additional and formally unrelated tool borrowed from
ecology, which we refer to as the Chao-Shen estimator. The predictions
of this estimator confirm that the relationship between sample size and
segment type count is best described by a curve that approximates
Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did
not converge on any single number of phoneme types as number of
languages increased.

Assuming a target of 7151 known languages,\footnote{\url{https://www.ethnologue.com/guides/how-many-languages}?}
our models predict a range of total segment types from 4592 to 4850
(mean = 4721). Based on PHOIBLE, our current best guess of the diversity
of documented phoneme types is 3183. Therefore, the models argue for the
existence of approximately 1500 yet to be documented phoneme types. As
we probe increasingly larger and theoretically motivated language sample
sizes, i.e., languages that existed \emph{through time} (e.g., estimates
including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
Herdan-Heaps' model predicted a continual increase in the number of
unique phonemes, reaching values as high as 19K. The Chao-Shen model
produced much more conservative estimates for samples of size 10K and
above, even appearing to settle on an asymptote of roughly 5500 phonemes
for samples of 100K languages or more. However, when we compare the
predictions of Herdan-Heaps and Chao-Shen against known type counts in a
closed dataset, the source of the discrepancy emerged: as the sample
grows, Chao-Shen lags behind Herdan-Heaps in approximating the true
value. Thus, given a larger pool of training data, we expect Chao-Shen
to approximate more closely the estimates of Herdan-Heaps' law.

@@

However, most widely accepted feature sets contain features that are
specified at different levels of phonetic detail: some features are more
abstract (i.e., general similarities such as +/- syllabic), while others
are more concrete (i.e., closer to the articulatory reality, e.g.,
advanced tongue root). Some are grounded in pure articulatory or
acoustic phonetics, while others involve phonological properties of
language. And looming in the background of these discussions, the debate
of whether distinctive features are innate and universal or rather
learned and emergent continues. Nevertheless, in general, distinctive
feature theories of different theoretical stances propose around two
dozen features (\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume
2006}), which still has the potential to generate over 16 million
distinct feature vectors (2\^{}24). Current consensus aside, there does
not appear to be a principled limit to the degree of phonetic
specificity that can be encoded by a feature. Making finer and finer
contrasts may always produce more and more features. Thus, the exponent
in the Feature Bounding calculation can itself (by hypothesis) be
infinite, in line with the predictions of the models presented here, and
therefore with the notion of infinite possible phonemes. That said,
contemporary feature sets can capture phonological classes for place and
manner of articulation, which neatly classify the phonological patterns
and processes in and across languages (cf. Mielke
(\protect\hyperlink{ref-Mielke2008}{2008})).

@@

SM was funded by the Swiss National Science Foundation (Grant
No.~PCEFP1\_186841).

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

SM conceptualization. SM and NAL designed the study, interpreted the
results, and wrote the paper.
